Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-08 analysis/paper/paper.Rmd
done for the week; added concluding statement to intro
incorporated TBs edits to Intro

---
title: "Predation of Northern Pikeminnow on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight extant spring/summer Chinook salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. For the MPG to be considered viable, at least five of these eight populations should meet viability criteria set forth by @McElhany2000 and @ICTRT2007, and is necessary for recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and in year with sufficient abundance, support local fisheries that contribute to the local economies. All Upper Salmon River MPG populations have become depleted in recent decades. Declines in survival of juvenile Chinook salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, simplification of rivers resulting from the construction of railroads and agriculture, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of adult returns are further impacted by ocean and downriver harvests, altered ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the above human activities and reduced abundances, action agencies have invested substantial funds in recent years to rehabilitate tributary ecosystems in an attempt to improve juvenile survival and adult spawning conditions.

@@

One potentially important, but perhaps under-appreciated source of mortality on Chinook salmon is predation on emigrating juveniles by piscivorious fishes, including both native and non-native species. As an example,...

@@

Spring/summer Chinook salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics, termed downstream rearing (DSR) and natal reach rearing (NRR) by [@Copeland2014]. The DSR migrants leave the natal spawning area as subyearlings during June through November and typically winter in downstream, mainstem habitats prior to emigration to the ocean the following spring as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence prior to emigration to the ocean as smolts. The diversity of emigration tactics provides for increases population resiliency due to changes in the freshwater rearing and migration environments and buffers against catastrophic events.

The Deadwater Slough is the result of human activities and resulted from an alluvial fan that formed at the bottom of Deadwater following a storm event and resulting failure of a mining dam on Dump Creek. Additionally, the Deadwater Slough occurs in a reach of the Salmon River believed to be an important overwinter rearing area for DSR emigrants attempting to exhibit concealment behavior to survive winter prior to emigration to the ocean the following spring. Deadwater Slough contains little to no hydrological or structure features (i.e., little fish cover) and fine sediment or mud substrate; all conditions making concealment behavior difficult and providing little refuge from predation. The Deadwater reach of the Salmon River is also important in that juvenile DSR and NRR emigrant from all eight extant populations must pass through the area during ocean migration.

In recent years, concerns arose that predation on juvenile Chinook salmon at or near the Deadwater Slough was not inconsequential to Chinook salmon populations in the Upper Salmon River MPG, impacting both DSR migrants in the fall and NRR emigrants during the following spring. We hypothesize that decreases in survival (or apparent survival) of juvenile Chinook salmon observed by @Ackerman2018, and similar for sockeye salmon by @Axel2015, are largely, or at least partially, due to increased densities of piscivorous predators in the Deadwater Slough area. =Beginning in 2019, we set out to evaluate potential piscivorous fish predator populations in the Deadwater Slough and evaluate their potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook salmon. Our objectives for the study were four-fold:

1. Estimate the abundance (or relative abundance) of potential predators in the Deadwater Slough during the peaks of fall (DSR) and spring (NRR) juvenile emigrations.
2. Document predation on juvenile Chinook salmon during the emigration periods using gastric lavage.
3. Using a bioenergetics approach, attempt to estimate the number of juvenile Chinook salmon consumed during defined emigration periods for DSR and NRR juveniles.
4. Estimate potential impacts to Chinook salmon adult returns based on the estimate of juvenile Chinook salmon consumed during the defined emigration periods.

>>>>

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. At least five of these eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

@@

One potentially important, but perhaps under-appreciated source of mortality on Chinook salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. As an example,...

@@

Spring/summer Chinook Salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014]. The DSR migrants leave the natal spawning area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal spawning areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migratory tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency.

The Deadwater Slough is located in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., homogeneous channel with fine substrate) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

We hypothesize that increased densitied of piscivorous predators in the Deadwater Slough area may explain the reduced survival (or apparent survival) observed for juvenile Chinook Salmon [@Ackerman2018] and Sockeye Salmon [@Axel2015]. In this study, we estimated the abundance of a piscivorous fish predator population in the Deadwater Slough and their potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. Our objectives for the study were four-fold:

1. Estimate the abundance (or relative abundance) of potential predators in the Deadwater Slough during the peaks of fall (DSR) and spring (NRR) juvenile emigrations.
2. Document predation on juvenile Chinook Salmon during the emigration periods using gastric lavage.
3. Use a bioenergetics approach to estimate the total consumption of juvenile DSR and NRR Chinook salmon during defined emigration periods.
4. Quantify the potential impacts of estimated predation on juvenile to Chinook Salmon adult returns.

We follow with a discussion of the various assumptions that went into the mark-recapture and bioenergetics models and assessment of impacts to adult returns and how violations of some assumptions may affect overall results and inferences from the study.
